---
title: "Assessment"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Seasonal depression and antidepressant prescriptions

## Brrr: Cold, long and sad winter days

Coming from Peru to Scotland was not only a cultural shock but an environmental one, if you will. I had never experienced such strikingly beautiful but equally emotionally debilitating winters. In Scotland, daylight is precious despite its scarcity and temperatures above 2 degrees are to dream of. And here, is when I first heard the term 'Seasonal depression' being used the most at this time of the year, Autumn; whereas it's a passing comment from a friend, a sitcom joke, or a genuine Google search during those long, dark, and cold winter days when you feel extra homesick.\
Throughout my time in Scotland I have wondered if seasonal depression is more a myth than a reality but have never quite gathered more than qualitative data in the form of conversations with friends and acquaintances. Therefore, I have decided to quantify these verbal leads through this report.

## Investigation breakdown

The research question to be explored is:

> To what extent do seasonal variations relating to daylight hours and temperature affect antidepressant prescribing rates across Scottish NHS Health Boards?

### Variables

The data time frame to be used is from March 2024 to March 2025 to assess the potential cyclicality of results and make adequate comparison between seasons.

#### Season Categorisation

**Spring**: March, April, May\
**Summer**: June, July, August\
**Autumn**: September, October, November\
**Winter**: December, January, February\

#### Antidepressant Prescriptions

Raw data from multiple institutional websites include all medications prescribed by NHS Health Boards. To differentiate and include only Antidepressant prescriptions, only prescriptions with a BNF item code starting with **'0403'** will be used.

This report aims to be a multi-factorial exploration of the environmental and social dimensions of antidepressant prescription rates in Scotland given that it's analysis will focus on three distinct aspects as shown in the hypotheses.

### Hypotheses

This report puts forward an overarching hypothesis:

> **Null (*H~0~*)**: Antidepressant prescription rates do not have seasonal patterns.

> **Alternative (*H~a~*)**: Antidepressant prescription rates have seasonal patterns.

Where within *H~a~*, the following is proposed:

> **1a.** Prescription rates increase during shorter months with shorter daylight hours and colder temperatures (autumn and winter) and decrease during those with longer daylight hours and warmer temperature (spring and summer).

In the assumption that *H~a~* is true, this report proposed two further sub-hypotheses that further enrich the multi-factorial lens through which seasonal effects inlfuence prescription rates:

> > **I.** The latitude of prescribing NHS Health Boards further influences the seasonality of antidepressant prescription rates, the more northern a health board is, the heavier the impact of these seasonal patterns.

> > **II.** Areas with a higher socioeconomic deprivation index (SMID) will have a consistently higher benchmark antidepressant use regardless of season, making the impact of varying daylight hours and temperature be even greater.

## Introduction to Data

### Health Board Data

January to June 2024: <https://www.opendata.nhs.scot/dataset/84393984-14e9-4b0d-a797-b288db64d088/resource/f0df380b-3f9b-4536-bb87-569e189b727a/download/hb_pitc2024_01_06-1.csv>\

July to December 2024: <https://www.opendata.nhs.scot/dataset/84393984-14e9-4b0d-a797-b288db64d088/resource/f3b9f2e2-66c0-4310-9b8e-734781d2ed0a/download/hb_pitc2024_07_12-1.csv>\

January to June 2025: <https://www.opendata.nhs.scot/dataset/84393984-14e9-4b0d-a797-b288db64d088/resource/9de908b3-9c28-4cc3-aa32-72350a0579d1/download/hb_pitc2025_01_06.csv>

### Met Office Seasonal Data

North Scotland:\
Daylight hrs - <https://www.metoffice.gov.uk/pub/data/weather/uk/climate/datasets/Sunshine/ranked/Scotland_N.txt>\
Temperature - <https://www.metoffice.gov.uk/pub/data/weather/uk/climate/datasets/Tmean/date/Scotland_N.txt>\

East Scotland:\
Daylight hrs - <https://www.metoffice.gov.uk/pub/data/weather/uk/climate/datasets/Sunshine/ranked/Scotland_E.txt>\
Temperature - <https://www.metoffice.gov.uk/pub/data/weather/uk/climate/datasets/Tmean/date/Scotland_E.txt>\

West Scotland:\
Daylight hrs - <https://www.metoffice.gov.uk/pub/data/weather/uk/climate/datasets/Sunshine/ranked/Scotland_W.txt>\
Temperature - <https://www.metoffice.gov.uk/pub/data/weather/uk/climate/datasets/Tmean/date/Scotland_W.txt>\

### Scottish Index of Multiple Deprivation (SMID quintiles) Data

Taken from the Public Health Scotland website:\
<https://www.opendata.nhs.scot/dataset/78d41fa9-1a62-4f7b-9edb-3e8522a93378/resource/acade396-8430-4b34-895a-b3e757fa346e/download/simd2020v2_22062020.csv>

### NHS Health Board Total Population Data

Also taken from the Public Health Scotland website:\
<https://www.opendata.nhs.scot/dataset/gp-practice-populations>

## Methodology

1.  Load all the required libraries on RStudio

```{r}
library(tidyverse)
library(here) # for the upkeep of the directory structure
library(janitor) # for data cleaning
library(gt) # for table building

```

2.  Load all the Health Board Data (January 2024:June 2025) into RStudio and use the `janitor` package by using the `clean_names()` function to have uniform names throughout datasets

```{r}
prescr_jan_june_2024 <- read_csv("https://www.opendata.nhs.scot/dataset/84393984-14e9-4b0d-a797-b288db64d088/resource/f0df380b-3f9b-4536-bb87-569e189b727a/download/hb_pitc2024_01_06-1.csv") %>% 
  clean_names()

prescr_july_dec_2024 <- read_csv("https://www.opendata.nhs.scot/dataset/84393984-14e9-4b0d-a797-b288db64d088/resource/f3b9f2e2-66c0-4310-9b8e-734781d2ed0a/download/hb_pitc2024_07_12-1.csv") %>% 
  clean_names()

prescr_jan_june_2025 <- read_csv("https://www.opendata.nhs.scot/dataset/84393984-14e9-4b0d-a797-b288db64d088/resource/9de908b3-9c28-4cc3-aa32-72350a0579d1/download/hb_pitc2025_01_06.csv") %>% 
  clean_names()

```

3.  Now filter out by "bnf_item_code" to keep only those prescriptions that are antidepressants (codes starting with '0403') and that were prescribed within our selected time frame (March 2024 - March 2025). Make sure to maintain the following: hbt, bnf_item_description, number_of_paid_items, and paid_date_month. Repeat for all Health Board datasets.

```{r}
antidepr_march_june_2024 <- prescr_jan_june_2024 %>% 
  filter(str_detect(bnf_item_code, "^0403")) %>%
  filter(paid_date_month != "202401" & paid_date_month != "202402") %>%
  select(c(hbt,bnf_item_code,bnf_item_description,number_of_paid_items,paid_date_month)) %>% 
  group_by(hbt, paid_date_month) %>%
  summarise(number_of_items = sum(number_of_paid_items, na.rm = TRUE)) %>% 
  arrange(paid_date_month)

antidepr_july_dec_2024 <- prescr_july_dec_2024 %>% 
  filter(str_detect(bnf_item_code, "^0403")) %>%
  select(c(hbt,bnf_item_code,bnf_item_description,number_of_paid_items,paid_date_month)) %>% 
  group_by(hbt, paid_date_month) %>%
  summarise(number_of_items = sum(number_of_paid_items, na.rm = TRUE)) %>% 
  arrange(paid_date_month)

antidepr_jan_march_2025 <- prescr_jan_june_2025 %>% 
  filter(str_detect(bnf_item_code, "^0403")) %>%
  filter(paid_date_month != "202504" & paid_date_month != "202505" & paid_date_month != "202506") %>%
  select(c(hbt,bnf_item_code,bnf_item_description,number_of_paid_items,paid_date_month)) %>% 
  group_by(hbt, paid_date_month) %>%
  summarise(number_of_items = sum(number_of_paid_items, na.rm = TRUE)) %>% 
  arrange(paid_date_month)
```

4. Join the three antidepressant datasets to create a unified March 2024-March 2025 prescription data object.
```{r}
yearly_antidepr_data <- antidepr_march_june_2024 %>% 
full_join(antidepr_july_dec_2024)

yearly_antidepr_data <- yearly_antidepr_data %>% 
  full_join(antidepr_jan_march_2025)
```

